# 2. Logistic regression

One way to move on from linear regression is to consider settings where the dependent (target) variable is **discrete**. This opens a wide range of possibilities for modelling phenomena beyond the assumptions of continuity or normality.

**Logistic regression** is a powerful method that is well suited for predicting and classifying data by working with **probabilities**. It belongs to a large family of statistical models called **Generalized Linear Models** (GLM). An important special case that involves a **binary** target (taking only the values 0 or 1) is the most typical and popular form of logistic regression. 

We will learn the concept of **odds ratio** (OR), which helps to understand and interpret the estimated coefficients of a logistic regression model. We also take a brief look at **cross-validation**, an important principle and technique for assessing the performance of a statistical model with another data set, for example by splitting the data into a **training set** and a **testing set**.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data set information

The dataset is a result of high school student performance measured from two Portuguese schools, in relation to their alcohol consumption. More detailed information about the original dataset can be found here <https://archive.ics.uci.edu/ml/datasets/Student+Performance>.

```{r, echo=FALSE}
alc <- read.table("data/alc.csv", sep = ";", header = TRUE)
```

This dataset uses "school", "sex", "age", "address", "famsize", "Pstatus", "Medu", "Fedu", "Mjob", "Fjob", "reason", "nursery","internet" as (student) identifiers. The rest are personal attributes describing their life, performance and alcohol consumptions.

The dataset includes 382 records with the 35 variables comprised of the mentioned above. **high_use** is determined based on quantified value of **alc_use**. **alc_use** is calculated based on both weekday (**Dalc**) and weekend (**Walc**) alcohol consumption:
```{r}
str(alc)
```

### Hypotheses

Amongst the variables, student final grade (**G3**), **absences**, weekly study time (**studytime**) as well as whether or not student go out with friends (**goout**) are some of the major factors that might influence their alcohol consumption. The consumption is generally known for affecting human cognitive performance and sleep/wake cycles, therefore student **G3**, **absences** and **studytime** are possibly associated. Socializing habit such as going out with friends is culturally associated with alcohol usage, thus **goout** is another factor to be examined.

### Numerical and graphical overview

Here is a brief summary of the dataset:

```{r, echo=FALSE}
summary(alc)
```

Student final grade (**G3**) can range from 0 to 20, with the maximum value in the dataset being 18. **Absences** purely represent the number of absences at school (from 0 to 93). **goout** is a numerical value, which range from 1 - very low to 5 - very high. **studytime** is also numerical, with `1 - <2 hours`, `2 - 2 to 5 hours`, `3 - 5 to 10 hours`, `or 4 - >10 hours`.

The following tables give initial hint that support the aforementioned hypotheses, where high alcohol consumption is linked to worse school performance and a more social life style.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library("dplyr")
alc %>% group_by(high_use) %>% summarise(count = n(), mean_grade = mean(G3))
alc %>% group_by(high_use) %>% summarise(count = n(), mean_absences = mean(absences))
alc %>% group_by(high_use) %>% summarise(count = n(), mean_study_time = mean(studytime))
alc %>% group_by(high_use) %>% summarise(count = n(), mean_go_out = mean(goout))
```

The following box plots depict the data in a more detailed manner. Even with outliers considered, the data still pretty much in line with the initial hypotheses.

```{r, echo=FALSE}
library(ggplot2)

g1 <- ggplot(alc, aes(x = high_use, y = G3))
g1 + geom_boxplot() + xlab("High alcohol use") + ylab("Final grade")

g2 <- ggplot(alc, aes(x = high_use, y = absences))
g2 + geom_boxplot() + xlab("High alcohol use") + ylab("Absences")

g3 <- ggplot(alc, aes(x = high_use, y = studytime))
g3 + geom_boxplot() + xlab("High alcohol use") + ylab("Weekly study time")

g4 <- ggplot(alc, aes(x = high_use, y = goout))
g4 + geom_boxplot() + xlab("High alcohol use") + ylab("Going out with friends")
```

### Regression model

### Establishing the relationship

### Graphical model validation
